Normalizations are not what you think; Explicit Scale Simulation in ALDEx2

### 2024-05-16 App-note.Rmd
go3-edits

High-throughput sequencing (HTS) is a ubiquitous tool used to explore many biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell. Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps. Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., size, microbial load, or total gene expression) of the measured biological system is degraded or lost. Increasingly, researchers are turning to modified experimental protocols (e.g., DNA spike-ins or cell counting) in an attempt to recover the lost biological variation in scale; see for example [@Vandeputte:2017aa;@Props:2017aa]. However, these protocols often do not recover meaningful biological variation but instead recover scale variation at some intermediate step in the sample preparation protocol (e.g., at the stage where the DNA spike-in was added). In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.

Biological variation in scale can represent an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of these systems includes both relative change (composition) and absolute abundance (scale), yet current methods access only the compositional information.

To be more concrete, let \(\mathbf{Y}\) denote an \(D \times N\) matrix of sequence counts with elements \(\mathbf{Y}_{dn}\) denoting the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\). Let \(\mathbf{W}\) also represent a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true (as opposed to measured) amount of class \(d\) in the biological system from which sample \(n\) was obtained. Succinctly, \(\mathbf{Y}\) denotes the observed HTS data whereas \(\mathbf{W}\) denotes the true counts in the system under study. We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., totals) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\). The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data. For brevity and concreteness, we discuss ALDEx2 in its simplest form as a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but note that it can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see Fernandes et al. [-@fernandes:2013].

@@

Gierliński et al. [-@Gierlinski:2015aa] conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf1 gene knockout. This dataset has been used to argue that only a small number of replicates are needed to identity differentially abundant genes and makes the recommendation that some tools should be used in some conditions because each tool has different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout. Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes. Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa]].  Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.

>>>>

High-throughput sequencing (HTS) is a ubiquitous tool used to explore  biological phenomenon such as gene expression (single-cell sequencing, RNA-sequencing, meta-transcriptomics), microbial community composition (16S rRNA gene sequencing, shotgun metagenomics) and differential enzyme activity (selex, CRISPR killing). HTS proceeds by taking a sample from the environment, making a library, multiplexing (merging) multiple libraries together, and then applying a sample of the multiplexed library to the flow cell. Each of these steps can be though of as a sampling step: a pool of DNA that could be measured is sampled and only that subsampled DNA is carried over to subsequent steps. Within each sampling step, the connection between the size of the sampled DNA pool and the scale (e.g., complexity, size, microbial load, or total gene expression) of the measured biological system is degraded or lost. One strategy is to include DNA spike-ins or cell counting) to include information on the lost biological variation [@Vandeputte:2017aa;@Props:2017aa]. However, this only recovers variation at the stage where the DNA spike-in was added. In short, the disconnect between sample-to-sample variation in sequencing depth and biological variation in scale remains an outstanding challenge.

Biological variation in scale can represent an important unmeasured confounder in HTS analyses [@Lovell:2015]. For example, cells transformed by the cMyc oncogene have about 3 times the amount of mRNA and about twice the rRNA content than  non-transformed cells [@Nie:2012aa], and this dramatically skews transcriptome analysis [@Loven:2012aa]. In addition, wild-type and mutant strains of cell lines, yeast or bacteria often have different growth rates, which would affect our ability to identify truly differentially abundant genes [@Yoshikawa:2011aa]. As another example, the total bacterial load of the vaginal microbiome differs by 1-2 orders of magnitude in absolute abundance between the healthy and bacterial vaginosis states [@Zozaya:2010], and the composition between these states is dramatically different [@Ravel:2010;@Hummelen:2010]. Thus, a full description of these systems includes both relative change (composition) and complexity or size (scale), yet current methods access only the compositional information.

A sequence count observation is summarized in a  \(D \times N\) count matrix \(\mathbf{Y}\) where \(\mathbf{Y}_{dn}\) indicates the number of measured DNA molecules mapping to feature \(d\) (e.g., a taxon or gene) in sample \(n\). We want this observation to tell us about the true environment \(\mathbf{W}\) also represented by a \(D \times N\) matrix but with elements \(\mathbf{W}_{dn}\) representing the true as opposed to the measured values in the biological system from which sample \(n\) was obtained.

We can think of \(\mathbf{W}\) as consisting of two parts, the scale \(\mathbf{W}^{\perp}\) (e.g., complexity or total) and the composition \(\mathbf{W}^{\parallel}\) (i.e., proportions). That is, \(\mathbf{W}^{\perp}\) is a \(N\)-vector with elements \(\mathbf{W}^{\perp}_{n}=\sum_{d}\mathbf{W}_{dn}\) while \(\mathbf{W}^{\parallel}\) is a \(D \times N\) matrix with elements \(\mathbf{W}^{\parallel}_{dn}=\mathbf{W}_{dn}/\mathbf{W}^{\perp}_{n}\). Note that with these definitions \(\mathbf{W}\) can be written as the element-wise combination of scale and composition: \(\mathbf{W}_{dn}=\mathbf{W}^{\parallel}_{dn}\mathbf{W}^{\perp}_{n}\). The technical variation in sequencing depth (\(\mathbf{Y}^{\perp}_{n}=\sum_{d}\mathbf{Y}_{dn}\)) implies that observed data \(\mathbf{Y}\) provides us with information about the system composition \(\mathbf{W}^{\parallel}\) but little to no information in the system scale \(\mathbf{W}^{\perp}\) (Lovell et al. 2011).

@@

The ALDEx2 R package [@fernandes:2013] is a general purpose toolbox for Bayesian modeling of HTS data. In its simplest form it is a tool for estimating Log-Fold-Changes (LFC) (e.g., differential abundance or differential expression analysis), but can be used to fit more complex linear models. At a high-level, ALDEx2 involves three steps: 1) a Bayesian model is used to estimate \(\mathbf{W}^{\parallel}\) given observations \(\mathbf{Y}\); 2) a normalization is used to estimate \(\mathbf{W}\) given the estimate \(\mathbf{W}^{\parallel}\); 3) the \(\mathbf{W}\) estimates are used to estimate LFC as part of differential abundance (or expression) analyses. For more details on ALDEx2 see Fernandes et al. [-@fernandes:2013].

@@

Gierliński et al. [-@Gierlinski:2015aa] conducted a highly replicated yeast transcriptome experiment comparing a wild-type strain with a snf1 gene knockout. This dataset has been used to argue that only a small number of replicates are needed to identity differentially abundant genes and makes the recommendation that some tools should be used in some conditions because each tool has different intrinsic statistical power and Type 1 and Type 2 errors [@Schurch:2016aa]]. This guidance runs counter to standard statistical practice where power is intrinsically linked to sample size [@Halsey:2015aa], yet the concept of sample-size independent power is entrenched in all fields that use HTS as an experimental readout. Through the lens of this paper, this is a hallmark of unacknowledged bias: a false certainty in the scale of the data brings about an increasing false precision with increasing sample sizes. Furthermore, the yeast transcriptome experiment has an unacknowledged difference in scale between conditions: yeast deficient for snf1 are smaller, grow more slowly and are sensitive to a variety of common agents that cause cell stress [@Yoshikawa:2011aa].  Thus, by adding scale uncertainty, we can be more confident that the results of the analysis reflect true biological differences and not differences due to the assumptions made by the normalization method.

### 2024-05-16 go3-author.tex
go3-edits

\begin{abstract}
In many high-throughput sequencing (HTS) studies, sample-to-sample
variation in sequencing depth is not driven by variation in the scale
(e.g., total size, microbial load, or total gene expression) of the
underlying biological systems being measured but rather is driven by
technical factors. Typically the technical variation is addressed using
some form of statistical normalization. Normalizations are data or
parameter transformations that remove unwanted technical variation in
the hopes of facilitating analyses sensitive to scale; e.g.,
differential abundance and differential expression analyses. Recently we
showed that any normalization makes implicit assumptions about the
unmeasured system scale and that errors in these assumptions can lead to
dramatic increases in false positive and false negative rates. Here we
describe updates to the ALDEx2 R package that mitigate these problems by
directly modeling uncertainty in the unmeasured system scale through the
use of a \textit{scale model}. Scale models generalize the idea of
normalizations and can be thought of as explicitly modeling potential
error in the chosen normalization. Beyond enhancing the robustness of
HTS analyses, the use of scale models within ALDEx2 enhances the
transparency and reproducibility of analyses by making implicit
normalizing assumptions an explicit part of the model building process.
\end{abstract}

@@

Gierliński et al. (2015) conducted a highly replicated yeast
transcriptome experiment comparing a wild-type strain with a snf1 gene
knockout. This dataset has been used to argue that only a small number
of replicates are needed to identity differentially abundant genes and
makes the recommendation that some tools should be used in some
conditions because each tool has different intrinsic statistical power
and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
runs counter to standard statistical practice where power is
intrinsically linked to sample size (Halsey et al. 2015), yet the
concept of sample-size independent power is entrenched in all fields
that use HTS as an experimental readout. Through the lens of this paper,
this is a hallmark of unacknowledged bias: a false certainty in the
scale of the data brings about an increasing false precision with
increasing sample sizes. Furthermore, the yeast transcriptome experiment
has an unacknowledged difference in scale between conditions: yeast
deficient for snf1 are smaller, grow more slowly and are sensitive to a
variety of common agents that cause cell stress (Yoshikawa et al.
2011){]}. Thus, by adding scale uncertainty, we can be more confident
that the results of the analysis reflect true biological differences and
not differences due to the assumptions made by the normalization method.

>>>>

\section{Abstract}\label{abstract}
In many high-throughput sequencing (HTS) studies, sample-to-sample
variation in sequencing depth is not driven by variation in the scale
(e.g., total size, microbial load, or total gene expression) of the
underlying biological systems being measured but rather is driven by
technical factors. Typically the technical variation is addressed using
some form of statistical normalization. Normalizations are data or
parameter transformations that remove unwanted technical variation in
the hopes of facilitating analyses sensitive to scale; e.g.,
differential abundance and differential expression analyses. Recently we
showed that any normalization makes implicit assumptions about the
unmeasured system scale and that errors in these assumptions can lead to
dramatic increases in false positive and false negative rates. Here we
describe updates to the ALDEx2 R package that mitigate these problems by
directly modeling uncertainty in the unmeasured system scale through the
use of a \textit{scale model}. Scale models generalize the idea of
normalizations and can be thought of as explicitly modeling potential
error in normalization by examining a distribution over all possible 
normalizations.  Beyond enhancing the robustness of HTS analyses, 
the use of scale models within ALDEx2 enhances the transparency and 
reproducibility of analyses by making implicit normalizing assumptions an 
explicit part of the model building process.
%\end{abstract}

@@

Gierliński et al. (2015) conducted a highly replicated yeast
transcriptome experiment comparing a wild-type strain with a snf1 gene
knockout. This dataset has been used to argue that only a small number
of replicates are needed to identity differentially abundant genes and
makes the recommendation that some tools should be used in some
conditions because each tool has different intrinsic statistical power
and Type 1 and Type 2 errors (Schurch et al. 2016){]}. This guidance
runs counter to standard statistical practice where power is
intrinsically linked to sample size (Halsey et al. 2015), yet the
concept of sample-size independent power is entrenched in all fields
that use HTS as an experimental readout. Through the lens of this paper,
this is a hallmark of unacknowledged bias: a false certainty in the
scale of the data brings about an increasing false precision with
increasing sample sizes. Furthermore, the yeast transcriptome experiment
has an unacknowledged difference in scale between conditions: yeast
deficient for snf1 are smaller, grow more slowly and are sensitive to a
variety of common agents that cause cell stress (Yoshikawa et al. 2011).
Thus, by adding scale uncertainty, we can be more confident that the
results of the analysis reflect true biological differences and not
differences due to the assumptions made by the normalization method.
